---
title: "R Notebook"
output: html_notebook
---

## VERİ MANİPULASYONU

Veriyi amaca yönelik olarak okunabilir ve düzenli hale getirmek

### Tidy Tools Manifestosu 4 Prensibi

- [ ] Mümkün olduğu kadar var olan veri yapılarını kullan
- [ ] Pipe ile basit fonksiyonlar oluştur: Büyük problemleri küçük parçalara ayırarak çöz
- [ ] Fonksiyonel programlamayı benimse
- [ ] İnsanlar için tasarla: Temiz ve okunabilir kod yaz

### Pipe Operatörü

%>%

## dplyr nedir?

A Grammar of Data Manipulation

### Cognitive Process

1. Take the ydat dataset, then
2. filter() for genes in the leucine biosynthesis pathway, then
3. group_by() the limiting nutrient, then
4. summarize() to correlate rate and expression, then
5. mutate() to round r to two digits, then
6. arrenge() by rounded correlation coefficients.

### The Old Way

```{r}

arrange(
  mutate(
    summarize(
      group_by(
        filter(ydat, bp=="leucine biosynthesis"),
        nutrient),
      r=cor(rate, expression)),
    r=round(r,2)),
  r)

```


### The dplyr Way

```{r}

ydat %>%
  
  filter(bp==“leucine biosynthesis”) %>%
  group_by(nutrient) %>%
  summarise(r=cor(rate, expression)) %>%
  mutate(r=round(r,2)) %>%
  arrange(r)

```

## %>% Example

### In Base R:

```{r}

# new table
gtbl1 = gtbl[gtbl$continent == "Asia",]

#reassignment
gtbl2 = gtbl1[gtbl1$lifeExp < 65,]

# functions in brackets
gtbl3 = gtbl2[sample(1:dim(gtbl2)[1], size = 10),]

gtbl3

```


### With pipes + dplyr

```{r}

gtbl %>%
  
  filter(continent == "Asia" & lifeExp < 65) %>%
  sample_n(10)
```


### Getting complicated, but readable

```{r}

gtbl %>%
  
  filter(year > 1960) %>%
  group_by(country) %>%
  sample_n(10) %>%
  mutate(pop_exp = pop / lifeExp,
         sampled = TRUE)
```


## Tidy Data - tidyr

variables : sutunlar
observations : satirlar
values : sutun ve satirdaki gözlemlerin hepsi!

